---
title: "Getting Started with RPPAnalyzeR"
subtitle: "RPPA Time-Course Analysis: MCF7 Serum Starvation Example"
author: "Amar Kumar"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with RPPAnalyzeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 9,
  fig.height = 6,
  dpi        = 150,
  warning    = FALSE,
  message    = FALSE
)
```

## Overview

`RPPAnalyzeR` provides a complete analysis pipeline for **Reverse Phase Protein Array (RPPA)** data produced by the [MD Anderson Cancer Center Functional Proteomics RPPA Core](https://www.mdanderson.org/research/research-resources/core-facilities/functional-proteomics-rppa-core.html).

This vignette walks through the full analysis of a serum-starvation time-course experiment in **MCF7 human breast cancer cells**, probed with **497 antibodies** across **5 time points** (0h, 2h, 4h, 8h, 24h).

---

## Installation

```r
# Install from GitHub
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("yourusername/RPPAnalyzeR")
```

Install required dependencies:

```r
install.packages(c(
  "readxl", "dplyr", "tidyr", "ggplot2", "pheatmap",
  "stringr", "tibble", "purrr", "openxlsx",
  "ggrepel", "RColorBrewer", "scales"
))
```

---

## Quickstart: One-Line Pipeline

The fastest path to results — runs everything and writes outputs to a folder:

```r
library(RPPAnalyzeR)

results <- run_pipeline(
  xlsx_path  = "01_Jonathan_Coloff__Vipin_Rawat.xlsx",
  output_dir = "rppa_analysis"
)
```

This produces:
- `rppa_analysis/RPPAnalyzeR_results.xlsx` — all key sheets in one workbook
- `rppa_analysis/DE_significant_hits.csv` — significant proteins across all time points
- `rppa_analysis/plots/` — 5+ publication-ready figures

---

## Step-by-Step Walkthrough

### 1. Import the RPPA Workbook

```r
library(RPPAnalyzeR)

rppa <- import_rppa("01_Jonathan_Coloff__Vipin_Rawat.xlsx")

# What did we get?
names(rppa)
#> [1] "l4_log2"        "l4_linear"      "l4_chm"
#>     "l3_log2"        "l2_log2"        "chm_timepoints"
#>     "antibody_qc"    "sample_qc"      "metadata"
#>     "sample_info"

# Check sample info
rppa$sample_info
```

The import reads all 12 sheets and organises them into a named list. The most important sheet is `l4_log2` — fully normalised (loading + batch) log2 expression values.

### 2. Quality Control

```r
rppa <- run_qc(
  rppa,
  min_qc_score       = 0.8,    # MD Anderson minimum threshold
  remove_caution     = FALSE,  # keep "C" antibodies (but they're flagged)
  remove_low_protein = TRUE    # exclude samples with Total Protein < -3
)

print_qc_summary(rppa)
```

**What the MD Anderson QC scores mean:**
- `0.8–1.0` → reliable data; 0.8 is the minimum acceptable
- Scores consider: signal strength, background noise, dilution curve quality
- All 497 antibodies in this dataset passed (scores 0.81–0.96+)
- All 15 samples passed total protein QC (none below -3 threshold)

Visualise the QC:

```r
plot_sample_qc(rppa)    # bar chart of total protein per sample
plot_antibody_qc(rppa)  # histogram of QC scores
```

### 3. Which Sheets Are Most Useful?

| Sheet | Use case |
|---|---|
| `rppa$l4_log2` | **Primary analysis** — normalised log2 values |
| `rppa$l4_chm` | **Heatmaps** — median-centred, matches the PDF heatmaps |
| `rppa$chm_timepoints` | **Pairwise comparisons** — each time point vs. 0h with Log2FC + P values |
| `rppa$antibody_qc` | Filter low-quality antibodies |
| `rppa$sample_qc` | Flag problematic samples |
| `rppa$metadata` | Antibody gene names, origin (R/M/G), validation status (V/C) |

```r
# Access the 0h vs 24h comparison directly
rppa$chm_timepoints$o_vs_24h %>%
  dplyr::filter(Significant) %>%
  dplyr::arrange(Pvalue) %>%
  head(10)
```

### 4. Differential Expression

```r
# One time point at a time
de_2h <- diff_expression(rppa, timepoint = "2h")

# All time points in one call
de_all <- diff_expression_all(rppa)

# Top 10 most significant proteins at 24h
top_proteins(de_all, n = 10, timepoint = "24h")
```

The CHM sheets in the Excel file already contain **t-test p-values** computed by Excel's TTEST function. `diff_expression()` reads these directly (via `use_chm_sheet = TRUE`) for consistency with the MD Anderson report. Setting `use_chm_sheet = FALSE` runs a fresh t-test from the L4 log2 values.

### 5. Visualizations

#### Volcano plot
```r
plot_volcano(de_all, timepoint = "24h", label_top = 15)
```

#### Clustered heatmap (top 50 most variable proteins)
```r
plot_heatmap(rppa, title = "Top 50 Proteins — MCF7 Serum Starvation")
```

#### Time-course for specific proteins
```r
# mTOR pathway proteins known to change with serum starvation
plot_timecourse(rppa, proteins = c(
  "p70-S6K_pT389",   # key mTOR readout
  "4E-BP1",          # mTOR cap-dependent translation
  "Akt_pS473",       # PI3K/Akt
  "MAPK_pT202_Y204", # ERK
  "S6_pS240_S244"    # ribosomal S6
))
```

### 6. Export Results

```r
# Full Excel workbook with all important sheets
export_rppa_excel(
  rppa,
  de_results  = de_all,
  output_path = "my_rppa_analysis.xlsx"
)

# Just the significant hits as CSV
export_de_csv(de_all, significant_only = TRUE,
              output_path = "significant_proteins.csv")
```

The exported Excel file contains:
- **L4_log2** — primary normalised data
- **L4_CHM** — heatmap-ready data
- **CHM_0h_vs_2h / 4h / 8h / 24h** — pairwise comparisons (highlighted: green = down, red = up)
- **DE_All_Timepoints** — combined differential expression
- **Antibody_QC** and **Sample_QC** — quality tables
- **Antibody_Metadata** — gene names, species, validation status

---

## Understanding the Data Levels

The MD Anderson pipeline produces 4 normalisation levels. RPPAnalyzeR imports all of them:

```
L2 (log2)  → Raw RPPA signal from RPPASPACE. No normalisation.
    ↓
L3 (log2)  → L2 + bidirectional median centering (loading normalisation).
    ↓
L4 (log2)  → L3 + batch effect correction (Set-to-Set adjustment).
    ↓
L4 (CHM)   → L4 + median centering by antibody (for heatmaps).
```

**Always use L4 (log2) for analysis.** L2 and L3 are included for transparency and troubleshooting, not for biological conclusions.

---

## Antibody Labels Decoded

In the heatmaps, antibody names follow the pattern: `AntibodyName-Species-Validation`

| Code | Meaning |
|---|---|
| **R** | Rabbit antibody |
| **M** | Mouse antibody |
| **G** | Goat antibody |
| **V** | Validated — performs well in all assays |
| **C** | Caution — good under most conditions, some discrepancies |
| **E** | Under evaluation |

Example: `Akt_pS473-R-V` = Akt phospho-Ser473, rabbit antibody, validated.

---

## Citation

If you use RPPAnalyzeR in your work, please cite:

1. The MD Anderson RPPA Core: Grant # 5 P30 CA016672-40
2. This package: [GitHub link]

---

## Session Info

```{r session}
sessionInfo()
```
