---
title: "RPPA Analysis Report"
subtitle: "MCF7 Serum Starvation Time-Course | Set208 | MD Anderson RPPA Core"
author: "Amar Kumar"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 10
    fig_height: 6
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  dpi       = 150
)

# ── USER: update this path to your Excel file ──────────────────────────────
xlsx_path <- "/Users/AMARMAC/Downloads/RPPAnalyzeR/01_Jonathan_Coloff__Vipin_Rawat.xlsx"

# ── USER: update this to where you want outputs saved ──────────────────────
output_dir <- "/Users/AMARMAC/Downloads/RPPAnalyzeR/my_results"

dir.create(file.path(output_dir, "plots"), recursive = TRUE, showWarnings = FALSE)
```

---

# Overview

This report documents the complete analysis of a **Reverse Phase Protein Array (RPPA)** dataset produced by the [MD Anderson Functional Proteomics RPPA Core](https://www.mdanderson.org/research/research-resources/core-facilities/functional-proteomics-rppa-core.html).

| Parameter | Value |
|---|---|
| Cell line | MCF7 human breast cancer cells |
| Experiment | Serum starvation time-course |
| Time points | 0h, 2h, 4h, 8h, 24h (3 replicates each) |
| Total samples | 15 |
| Antibodies probed | 497 |
| RPPA Set | Set208, May 2025 |
| Data level used | L4 (log2) — loading + batch normalised |

---

# 1. Setup

```{r load_package}
library(RPPAnalyzeR)
library(dplyr)
library(ggplot2)
library(knitr)
```

---

# 2. Import Data

All 12 sheets from the MD Anderson Excel workbook are imported in one call. The disorganised 9-row metadata header is parsed automatically.

```{r import}
rppa <- import_rppa(xlsx_path, verbose = TRUE)
```

## 2.1 What was imported

```{r import_summary}
summary_tbl <- data.frame(
  Sheet = c("L4 (log2)", "L4 (linear)", "L4 (CHM)",
            "L3 (log2)", "L2 (log2)",
            "CHM time-point sheets", "Antibody QC", "Sample QC", "Metadata"),
  Contents = c(
    "Primary data — fully normalised (loading + batch)",
    "L4 converted to linear scale — for bar graphs",
    "L4 antibody-median-centred — for heatmaps",
    "Loading-normalised only (intermediate)",
    "Raw unnormalised RPPASPACE output",
    paste0(length(rppa$chm_timepoints), " pairwise comparisons vs 0h with Log2FC and p-values"),
    paste0(nrow(rppa$antibody_qc), " antibodies with QC scores"),
    paste0(nrow(rppa$sample_qc), " samples with total protein values"),
    paste0(nrow(rppa$metadata), " antibodies with gene names, species, validation status")
  )
)
kable(summary_tbl, caption = "Sheets imported from RPPA Core Excel file")
```

## 2.2 Sample overview

```{r sample_table}
rppa$sample_info %>%
  select(Sample_Name, Sample_Description, Sample_Type, Timepoint, Replicate) %>%
  kable(caption = "All 15 samples in this dataset")
```

## 2.3 Antibody metadata (first 20)

```{r metadata_preview}
rppa$metadata %>%
  head(20) %>%
  select(Antibody_Name, Gene_Name, Antibody_Origin, Validation_Status) %>%
  kable(caption = "Antibody metadata — first 20 of 497")
```

---

# 3. Quality Control

## 3.1 Sample QC — Total Protein Content

Total protein content reflects the average signal of a sample across all antibodies. Values below log2 = −3 are flagged as unreliable by MD Anderson and should be excluded.

```{r sample_qc_plot, fig.height=5, fig.cap="Total protein content per sample. Red dashed line = exclusion threshold (log2 < -3)."}
p_sqc <- plot_sample_qc(rppa)
print(p_sqc)
save_plot(p_sqc, file.path(output_dir, "plots", "sample_QC.png"), width = 10, height = 5)
```

```{r sample_qc_table}
rppa$sample_qc %>%
  select(Sample_Description, Timepoint, Total_Protein, Low_Protein) %>%
  mutate(
    Total_Protein = round(Total_Protein, 3),
    Status        = ifelse(Low_Protein, "FAIL — exclude", "PASS")
  ) %>%
  select(-Low_Protein) %>%
  kable(caption = "Sample QC metrics")
```

## 3.2 Antibody QC — Slide Staining Scores

QC scores range from 0 (worst) to 1 (best). MD Anderson requires a minimum score of 0.8 for inclusion.

```{r antibody_qc_plot, fig.height=5, fig.cap="Distribution of antibody QC scores. Red dashed line = minimum threshold (0.8)."}
p_aqc <- plot_antibody_qc(rppa)
print(p_aqc)
save_plot(p_aqc, file.path(output_dir, "plots", "antibody_QC.png"), width = 8, height = 5)
```

## 3.3 Apply QC filters

```{r run_qc}
rppa <- run_qc(
  rppa,
  min_qc_score       = 0.8,
  remove_caution     = FALSE,
  remove_low_protein = TRUE
)

print_qc_summary(rppa)
```

```{r qc_report_table}
qr <- rppa$qc_report
data.frame(
  Parameter = c(
    "Antibodies input", "Antibodies kept", "Antibodies removed (low QC)",
    "Caution (C) antibodies retained",
    "Samples input", "Samples kept", "Low-protein samples removed"
  ),
  Value = c(
    qr$n_antibodies_input, qr$n_antibodies_kept, qr$n_antibodies_removed,
    length(qr$caution_antibodies),
    qr$n_samples_input, qr$n_samples_kept,
    length(qr$low_protein_samples)
  )
) %>% kable(caption = "QC filter summary")
```

---

# 4. Differential Expression Analysis

For each time point, proteins are compared against the 0h baseline using the pre-computed t-test p-values from the MD Anderson CHM sheets. Significance threshold: **|Log2FC| > 0.5 and P < 0.05**.

```{r run_de}
de_all <- diff_expression_all(
  rppa,
  ctrl_timepoint = "0h",
  p_threshold    = 0.05,
  fc_threshold   = 0.5
)
```

## 4.1 Summary of significant hits per time point

```{r de_summary_table}
de_all %>%
  group_by(Timepoint) %>%
  summarise(
    Total_tested  = n(),
    Significant   = sum(Significant, na.rm = TRUE),
    Upregulated   = sum(Direction == "Up",   na.rm = TRUE),
    Downregulated = sum(Direction == "Down", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  kable(caption = "Significant proteins at each time point vs 0h baseline")
```

## 4.2 Top 15 proteins at each time point {.tabset}

### 2h vs 0h
```{r top_2h}
top_proteins(de_all, n = 15, timepoint = "2h") %>%
  select(Protein, Mean_Ctrl, Mean_Treat, Log2FC, Pvalue, Direction) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Top 15 significant proteins at 2h")
```

### 4h vs 0h
```{r top_4h}
top_proteins(de_all, n = 15, timepoint = "4h") %>%
  select(Protein, Mean_Ctrl, Mean_Treat, Log2FC, Pvalue, Direction) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Top 15 significant proteins at 4h")
```

### 8h vs 0h
```{r top_8h}
top_proteins(de_all, n = 15, timepoint = "8h") %>%
  select(Protein, Mean_Ctrl, Mean_Treat, Log2FC, Pvalue, Direction) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Top 15 significant proteins at 8h")
```

### 24h vs 0h
```{r top_24h}
top_proteins(de_all, n = 15, timepoint = "24h") %>%
  select(Protein, Mean_Ctrl, Mean_Treat, Log2FC, Pvalue, Direction) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Top 15 significant proteins at 24h")
```

---

# 5. Volcano Plots {.tabset}

Each point is one protein. Red = significantly upregulated, green = significantly downregulated, grey = not significant. Labels show the top 15 most significant proteins.

### 2h vs 0h
```{r volcano_2h, fig.height=7, fig.cap="Volcano plot: 2h vs 0h baseline"}
p_v2 <- plot_volcano(de_all, timepoint = "2h", label_top = 15)
print(p_v2)
save_plot(p_v2, file.path(output_dir, "plots", "volcano_0h_vs_2h.png"), width = 8, height = 7)
```

### 4h vs 0h
```{r volcano_4h, fig.height=7, fig.cap="Volcano plot: 4h vs 0h baseline"}
p_v4 <- plot_volcano(de_all, timepoint = "4h", label_top = 15)
print(p_v4)
save_plot(p_v4, file.path(output_dir, "plots", "volcano_0h_vs_4h.png"), width = 8, height = 7)
```

### 8h vs 0h
```{r volcano_8h, fig.height=7, fig.cap="Volcano plot: 8h vs 0h baseline"}
p_v8 <- plot_volcano(de_all, timepoint = "8h", label_top = 15)
print(p_v8)
save_plot(p_v8, file.path(output_dir, "plots", "volcano_0h_vs_8h.png"), width = 8, height = 7)
```

### 24h vs 0h
```{r volcano_24h, fig.height=7, fig.cap="Volcano plot: 24h vs 0h baseline"}
p_v24 <- plot_volcano(de_all, timepoint = "24h", label_top = 15)
print(p_v24)
save_plot(p_v24, file.path(output_dir, "plots", "volcano_0h_vs_24h.png"), width = 8, height = 7)
```

---

# 6. Clustered Heatmap

Top 50 most variable proteins across all samples, using L4 CHM (antibody-median-centred) values. Red = high expression, green = low expression. Samples are kept in time-point order; proteins are clustered by similarity.

```{r heatmap, fig.height=14, fig.cap="Clustered heatmap of top 50 most variable proteins"}
# Get top 50 most variable protein names
expr  <- get_expression_matrix(rppa, "l4_chm")
vars  <- apply(expr$matrix, 2, var, na.rm = TRUE)
top50 <- names(sort(vars, decreasing = TRUE))[1:50]
top50_clean <- stringr::str_remove(top50, "-[RMG]-[VCE]$")

# Draw the heatmap inline in this report
plot_heatmap(
  rppa,
  proteins = top50_clean,
  title    = "Top 50 Most Variable Proteins — MCF7 Serum Starvation (L4 CHM)"
)

# Save a copy to plots/. Note: pheatmap writes to file *instead of* drawing when
# `filename` is supplied, so the call above is needed for the inline figure and
# this second call is needed for the saved PNG.
plot_heatmap(
  rppa,
  proteins = top50_clean,
  title    = "Top 50 Most Variable Proteins — MCF7 Serum Starvation (L4 CHM)",
  filename = file.path(output_dir, "plots", "04_heatmap_top50.png")
)
```

---

# 7. Time-Course Plots {.tabset}

Mean ± SE expression across replicates at each time point (L4 log2 values).

### mTOR / PI3K pathway
```{r timecourse_mtor, fig.cap="mTOR and PI3K pathway proteins over time"}
p_tc1 <- plot_timecourse(rppa, proteins = c(
  "p70-S6K_pT389-R-V",
  "Akt_pS473-R-V",
  "4E-BP1-R-V",
  "mTOR_pS2448-R-C",
  "S6_pS240_S244-R-V",
  "PRAS40_pT246-R-V"
))
print(p_tc1)
save_plot(p_tc1, file.path(output_dir, "plots", "timecourse_mTOR_PI3K.png"), width = 9, height = 6)
```

### MAPK / ERK pathway
```{r timecourse_mapk, fig.cap="MAPK and ERK pathway proteins over time"}
p_tc2 <- plot_timecourse(rppa, proteins = c(
  "MAPK_pT202_Y204-R-C",
  "MEK1_pS217_S221-R-V",
  "eEF2K-R-V",
  "RSK1-R-V",
  "Mnk1-R-V",
  "C-Raf_pS338-R-V"
))
print(p_tc2)
save_plot(p_tc2, file.path(output_dir, "plots", "timecourse_MAPK_ERK.png"), width = 9, height = 6)
```

### Cell cycle / apoptosis
```{r timecourse_cc, fig.cap="Cell cycle and apoptosis proteins over time"}
p_tc3 <- plot_timecourse(rppa, proteins = c(
  "Cyclin-B1-R-V",
  "Cyclin-D1-R-C",
  "p27-Kip1-R-C",
  "Rb_pS807_S811-R-V",
  "Caspase-3-cleaved-R-C",
  "PARP-R-V"
))
print(p_tc3)
save_plot(p_tc3, file.path(output_dir, "plots", "timecourse_CellCycle.png"), width = 9, height = 6)
```

---

# 8. Export Results

```{r export_excel}
export_rppa_excel(
  rppa,
  de_results  = de_all,
  output_path = file.path(output_dir, "RPPAnalyzeR_results.xlsx")
)
```

```{r export_csv}
export_de_csv(
  de_all,
  output_path      = file.path(output_dir, "significant_hits.csv"),
  significant_only = TRUE
)

n_sig <- sum(de_all$Significant, na.rm = TRUE)
cat("Total significant hits exported:", n_sig, "\n")
cat("Output folder:", output_dir, "\n")
```

---

# 9. Session Info

For reproducibility — records the exact R version and package versions used.

```{r session_info}
sessionInfo()
```

---

# Acknowledgements

RPPA data was generated by the **MD Anderson Functional Proteomics RPPA Core**.

> *The Functional Proteomics RPPA Core is supported by MD Anderson Cancer Center Support Grant # 5 P30 CA016672-40.*

---

*Report generated by RPPAnalyzeR v0.1.0 — Amar Kumar*
